Estimating Operator Intent\\ (Draft 3.0)

### 2015-09-28 control.tex
more updates

@@

We have not seen a lot of discussion of state estimation of exoskeletons in
the literature.
We will discuss our approach to state estimation on humanoid robots here.
State estimation to deal with sensor noise,
disturbance estimation, and incomplete or
redundant sensing is a separate process and is as independent as possible
from the control method~\cite{certainty-eq,separation-prin}.
This makes evaluating different control methods much easier.
Disturbance observers for disturbances such as
moving support surfaces, contact or wind forces, and model errors
are also implemented.

>>>>

This recent paper discusses shared control \url{http://www.ri.cmu.edu/pub_files/2015/7/Javdani15Hindsight.pdf}

@@

We use the operator state (the set of all joint angles and velocities
as well as a root position and velocity to represent operator location
and orientation in the world) as the basis of inverse dynamics and
online optimization-based control.
In a full-body exoskeleton
(not necessarily fully actuated), the
exoskeleton can be used to measure operator positions, and velocities.
For partial exoskeletons,
it may be the case that additional sensors such as distributed 
angle sensors (goniometers), tachometers, or inertial measurement units (IMUs)
are attached to the operator or part of the operator's clothing.
A major goal of state estimation is to process these measurements to
produce a useful estimate of operator state.

We have not seen a lot of discussion of state estimation of exoskeletons in
the literature.
In lieu of that,
we will discuss our approach to state estimation on humanoid robots here.
State estimation to deal with sensor noise,
disturbance estimation, and incomplete or
redundant sensing is a separate process and is as independent as possible
from the control method~\cite{certainty-eq,separation-prin}.
This makes evaluating different control methods much easier.
Disturbance observers for disturbances such as
moving support surfaces, contact or wind forces, and model errors
are also implemented.

### 2015-09-28 intent.tex
cga-fixes
cga updates
more updates

\newcommand{\gc}{{\mbox {GravityCompensation}}}
\newcommand{\invdyn}{{\mbox {\tiny ID}}}

@@

\title{Estimating Operator Intent\\
(Draft 2.1)}

@@

This paper surveys state of the art technology to recognize
and predict exoskeleton
operator intent.
A companion paper surveys current operator/exoskeleton interfaces~\cite{}.

@@

We focus this survey on intent estimation for lower body tasks (standing, walking,
running, jumping, kicking, dodging, ...).
We do not survey intent estimation for manipulation.

A future white paper will discuss how intent estimation could be 
made easier with superhuman perception (sensing beyond human capabilities
or modalities).

A future white paper will discuss how intent estimation could be 
made easier with implanted markers and devices, as well as continuous
operator imaging such as ultrasound.

Predictive models of human motion, a major topic of robot learning from\\
observation/demonstration/imitation and computer animation,
will be the subject of a future white
paper~\cite{IEEE06913830,Bagnell}

\section{Introduction: Our Point of View}

{\bf What is intent estimation?}
The further in the future the control system can predict what the operator
is going to do, the easier it is to control the exoskeleton. One way
to think about estimating operator intent is to imagine you are in
a hand-to-hand fight and you want to predict your opponent's next move.
You can also think about playing poker. What ``tells'' does the opponent
have?
You are allowed to instrument your opponent. What sensors would you use?
How would you interpret the sensors?

@@

{\bf Individualization of Intent Estimation:}
We recommend each exoskeleton controller
to be used by and optimized for a single operator.
A substantial investment in capturing the operators normal behavior,
operator training and learning, and controller customization should be made.

@@

\begin{verbatim}
10 sec - rely on perceiving situation and thinking like your opponent.
         What are the probabilities of various attacks? Responses?
1 sec - measure set/tells to estimate probability of particular behaviors.
100 msec - EMG, muscle state, force helpful here
10 msec - EMG, muscle state, force helpful here and
         immediate sensors: joint pos,vel, IMU, contact forces
1 msec - immediate sensors: joint pos,vel, IMU, contact forces
\end{verbatim}

@@

{\bf EMG:}
Electrical activity was measured using surface electrodes,
high-pass filtered at 20 Hz, rectified, low-pass filtered at 6 Hz,
offset by a small value of -0.008, and amplified by a gain of 283
to obtain the desired torque.~\cite{IEEE07139980}

Electromyography (EMG)
signals are extremely noisy and it is not easy to extract a
meaningful data form the raw signal. In addition, it takes a long
time to attach the EMG electrodes to the body and calibrate the
sensors.
\begin{verbatim}
T. Hayashi, H. Kawamoto, and Y. Sankai, "Control method of robot
suit HAL working as operator's muscle using biological and dynamical
information," in Intelligent Robots and Systems, 2005.(IROS 2005).
2005 IEEE/RSJ International Conference on, 2005, pp. 3063-3068.
H. Kawamoto and Y. Sankai, "Power assist system HAL-3 for gait
disorder person," in Computers helping people with special needs, ed:
Springer, 2002, pp. 196-203.
\end{verbatim}

Muscle movement leads to surface electrodes losing good contact, fully detaching, or
even ripping off (especially for large fast
movements, the worst time).
Perspiration changes measurement. EMG measurement
needs to be calibrated for each electrode application and periodically during use.

FMG-[06913842]

{\bf Foot contact and force/torque sensing:}
\begin{verbatim}
[8] K. Kong and M. Tomizuka, “A Gait Monitoring System Based on Air
Pressure Sensors Embedded in a Shoe,” IEEE/ASME Transactions on
Mechatronics, vol. 14, pp. 358–370, 2009.
\end{verbatim}

\section{How can the operator communicate with the exoskeleton}

\begin{verbatim}
Operator uses high frequency range or patterns of output to
signal what to do
 - selection
 - parameterization (how far, how high, how fast)
\end{verbatim}

The operator controls more than the pose and motion of the
exoskeleton. What if there are "pointing" sensors or communication
devices that need to be aimed at or track an area of interest? What if
there are additional (physical or virtual) pan/tilt/zoom cameras
pointing to the side and rear or line of sight secure communications,
for example. What about other controls, such as power assist level or
thermal control? How can an operator naturally express intent and
control these additional degress of freedom?

\section{Operator load and fatigue}

To what extent can task recognition and "autonomy" reduce operator
load and fatigue?

>>>>

@@

\title{Estimating Operator Intent\\
(Draft 3.0)}

@@

This paper surveys state of the art technology to recognize
and predict exoskeleton
operator intent.
A companion paper surveys current operator/exoskeleton interfaces [survey.pdf].

@@

We focus this survey on intent estimation for highly dynamic 
lower body tasks (standing, walking,
running, jumping, kicking, dodging, ...).
We do not survey intent estimation for manipulation.

A future white paper could discuss how intent estimation could be 
made easier with outward facing
superhuman perception (sensing beyond human capabilities
or modalities) such as cameras looking out all over the skin
of the suit (whole-body vision), or whole-body 
infrared vision of the situation.

A future white paper could discuss how intent estimation could be 
made easier with inward facing superhuman perception,
such as implanted markers and devices, as well as continuous
operator imaging such as reflective and/or transmissive ultrasound
tracking tissue parameters or implanted markers.

Predictive models of human motion, a major topic of robot learning from\\
observation/demonstration/imitation and computer animation,
could be the subject of a future white
paper~\cite{IEEE06913830}
\url{http://www.ri.cmu.edu/publication_view.html?pub_id=7891}

\section{Introduction: Our point of view}

{\bf What is intent estimation?}
The further in the future the control system can predict what the operator
is going to do, the easier it is to control the exoskeleton. One way
to think about estimating operator intent is to imagine you are in
a hand-to-hand fight (such as various forms of martial arts)
and you want to predict your opponent's next move.
You can also think about playing poker. What ``tells'' does the opponent
have?
You are allowed to instrument your opponent. What sensors would you use?
How would you interpret the sensors?

@@

\section{Who is the customer of intent predictions?}

One customer of intent estimation is an exoskeleton control system.
Other possible customers are online operator situation awareness,
cognitive assistance
and physical guidance,
command and squad visualization, situation awareness,
coordination, and cognitive assistance,
after-action reports, and new operator training and
operator learning from practice and coaching,
We will focus on exoskeleton control in this paper.

\section{What is the output of intent estimation?}

There are several possible outputs of intent estimation of use to the
exoskeleton control system.
Estimates and predictions could be actual values at the current time or at
future times, probability distributions for
values at various times such as Gaussians or multi-modal distributions 
such as mixtures of Gaussians, or probability distributions for trajectories
of values, supporting non-Markov models of processes such as libraries
of behaviors.

\subsection{Operator physical state}

It is useful to estimate the full state of the operator in terms
of joint positions and velocities.

\subsection{Operator-exoskeleton force}

One possible output of intent estimation is an estimate of current
and predicted future forces between the operator and the exoskeleton.

\subsection{Desired movement trajectory}

Exoskeleton model-based control systems based on inverse dynamics often
can make use of an estimate of the current desired acceleration.
Predictions can be in the forms of sequences of desired accelerations
across time, or some form of desired velocity or position trajectory,
such as a spline.

\subsection{Behavior recognition}

Forces and movement are continuous values. It may also be useful to
recognize and/or predict discrete behaviors, from a selection of
possible behaviors such as 
standing, walking,
running, jumping, kicking, and dodging.
It may be useful to maintain a probability distribution of possible
behaviors and reflexes at all times.

\subsection{Utility functions}

It may be useful to estimate and predict operator intent in the form
of criteria the operator is ``optimizing''. These functions have
various names in different fields: utility or cost functions in economics
and decision sciences,
reward functions in reinforcement learning, and
optimization criteria and one-step-cost functions in optimal control.
Estimated utility functions may have a predetermined form and unknown
parameters and weights.
Predicting optimization criteria is known as ``inverse optimal control''
in optimal control and reinforcement learning.
\url{http://www.ri.cmu.edu/publication_view.html?pub_id=7800}

\subsection{Operator mental state}

Estimating operator mental state: what they are attending to (sensing),
what they are thinking about (cognition), what they are aware of (short term
memory), and what
they know (long term memory) will improve estimation of operator intent.

@@

\begin{itemize}
\item
10 sec - {\bf Behavior selection.}
Rely on perceiving situation and thinking like your operator and your
opponents.
What are the probabilities of various attacks? Responses?
\item
1 sec - {\bf Behavior selection and reflex responses.}
Measure set/tells to estimate probability of particular behaviors and reflexes.
Maintain probability distribution of possible behaviors and reflexes.
\item
100 msec - {\bf Muscle activation.} EMG, muscle activation modeling
force helpful here.
\item
10 msec - {\bf Biomechanics.} Joint position, velocity, contact forces,
internal muscle and tissue forces.
\item
1 msec - {\bf Local and global accelerations and local deformation dynamics.}
Distributed MEMs IMUs on operator sensing local accelerations,
angular velocities, vibration, impacts, shock waves, and other high
frequency mechanical phenomenon.
\end{itemize}

@@

\subsection{What the operator sees, hears, feels, and smells}

We could have the control computer aware of the external world
available to the operator. These sensors are described in
a companion paper [technology.pdf].

\subsection{Brain signals}

It is possible that external measurements of brain activity
(electroencephalography (EEG)) could detect stimuli processing, and
even more usefully, operator decision making.
It is also possible that implanted measurement devices could improve
greatly over external measurement devices in terms of specificity,
timing, and reliability of brain signal measurements.
\url{https://en.wikipedia.org/wiki/Electroencephalography}
Measurement of brain signals,
in the context of a wide range of other measurements,
might improve overall intent estimation.

\subsection{EMG}

Muscle electrical activity can be measured by surface electrodes
(not very specific to particular muscles), needle electrodes (painful,
especially during movement), and implanted electrodes~\url{http://www.medscape.com/viewarticle/542356_2}.
Typical processing is to 
high-pass filter at 20 Hz, rectify, low-pass filter at 6 Hz,
offset by a small value, and amplify by some scaling factor
to get a muscle activation estimate~\cite{IEEE07139980}.
One can apply pattern recogntion to this signal~\cite{IEEE07222568}.

Electromyography 
signals are noisy and it is not easy to extract
meaningful data from the raw signal. 
EMG sensors
need to be attached to the operator, calibrated,
and periodically recalibrated during use, which takes time.
Perspiration changes the measurement. 
Muscle movement leads to surface electrodes losing good contact,
fully detaching, or
even ripping off (especially for large fast
movements, the worst time).

\subsection{Operator-exoskeleton force sensing}

Operator-exoskeleton forces may be measured (forcemyography (FMG))
with various types of
force sensors such as single and multi-axis strain-gage load cells,
force sensitive resistors (FSRs), and pressure sensors built into an internal exoskeleton skin or at the
contacts (straps and cushions) of the exoskeleton attachment points.
Optical sensors and other distance sensors can measure deformation and
strain of the operator and between the operator and the exoskeleton.
Overall muscle ``pressure'' and other tissue parameters can be measured using
elastic straps around the arm, deformation pressure probes at various
points on the arm.
An estimate of individual muscle activity
may be available from tissue impedance measurements on the arm
or through the arm (including various forms of reflective and transmissive
ultrasound imaging tracking tissue parameters or implanted markers).
Some or all operator-exoskeleton forces may not be measured, and must
be estimated with some sort of Kalman filter.
Muscle activation can be measured (electromyography (EMG))
and force dynamics can be estimated to predict future operator-exoskeleton
forces.

\subsection{Operator state and motion}

In a full-body exoskeleton
(not necessarily fully actuated), the
exoskeleton can be used to measure positions, velocities,
and accelerations. For partial exoskeletons,
it may be the case that additional sensors such as distributed 
angle sensors, tachometers, or inertial measurement units (IMUs)
are attached to the operator or part of the operator's clothing.

\section{Modeling the individual operator}

We recommend each exoskeleton controller
to be used by and optimized for a single operator.
A substantial investment in capturing the operators normal behavior,
operator training and learning, and controller customization should be made.
This paper discusses ways to model individual movement style
\url{http://humanmotion.ict.ac.cn/papers/2015P1_StyleTransfer/details.htm}
This paper discusses activity forecasting
\url{http://www.ri.cmu.edu/publication_view.html?pub_id=7250}

Predictive models of human motion, a major topic of robot learning from\\
observation/demonstration/imitation and computer animation,
could be the subject of a future white
paper~\cite{IEEE06913830}
\url{http://www.ri.cmu.edu/publication_view.html?pub_id=7891}

\section{Can the operator help?}

We have a cooperative operator, who can signal using unused frequency
ranges or patterns on existing measurement
channels (brain signals, EMG, force, and motion) or use 
task specific intefaces such as built-in keyboards, touch sensors, joysticks,
and speech recognition.
Operators can resolve ambiguities, select behaviors, and indicate
parameterization (how far, how high, how fast) using these interfaces.

We note that the operator will use more command channels that just
body movement, since
the operator controls more than the pose and motion of the
exoskeleton. What if there are ``pointing'' sensors or communication
devices that need to be aimed at or track an area of interest? What if
there are additional (physical or virtual) pan/tilt/zoom cameras
pointing to the side and rear or line of sight secure communications,
for example. What about other controls, such as power assist level or
thermal control? It is an open question
how an operator can naturally express intent and
control these additional degress of freedom?
